---
title: "Unit4 Homework"
author: "Richard Palmer"
date: "9/17/2019"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Conceptual Questions

### For a time series to be stationary

- Constant Mean
- Constant Variance
- Constant Autocorrelations

### if a timee series model ..
includes and explantory variable such as time itself, or some other predictor,
then the original time series of the response is not stationary.
(FALSE) - this could *hide* nonstationarity, but would not mean it is necessarily present.

### The main drawback of serially correlated observations is..

That there is less information from which to create a model.

### The main advantage of serrially correlated observations is..

that it is possible to make predictions about the future values of the time series.

### The purpose of the Durbin Watson test is...

to detect the presence of serial correlation in a time series.

## Exercise 1

6.  Verity rules of thumb

```{r}
AR1 <- arima.sim(list(ar=c(0.8)),10000)
par(mfrow=c(1,3))
plot(1:10000,AR1,type="l")
ar1_acf <- acf(AR1,main="ACF")
ar1_pacf <-pacf(AR1,main="PACF")
```

The acf clearly decays gradually.  The lag1 autocorrelation is `r unlist(ar1_acf)[2]`.

7.  Repeat 3 times with 50 observations each

```{r}
AR1_short = arima.sim(list(ar=c(0.8)),50)
par(mfrow=c(1,3))
plot(1:50,AR1_short,type="l")
AR1_short_acf <- acf(AR1_short,main="ACF")
AR1_short_pacf <-pacf(AR1_short,main="PACF")
```

The lag1 autocorrelation is `r unlist(AR1_short_acf)[2]`.

```{r}
AR1_short = arima.sim(list(ar=c(0.8)),50)
par(mfrow=c(1,3))
plot(1:50,AR1_short,type="l")
AR1_short_acf <- acf(AR1_short,main="ACF")
AR1_short_pacf <-pacf(AR1_short,main="PACF")
```

The lag1 autocorrelation is `r unlist(AR1_short_acf)[2]`.

```{r}
AR1_short = arima.sim(list(ar=c(0.8)),50)
par(mfrow=c(1,3))
plot(1:50,AR1_short,type="l")
AR1_short_acf <- acf(AR1_short,main="ACF")
AR1_short_pacf <-pacf(AR1_short,main="PACF")
```

The lag1 autocorrelation is `r unlist(AR1_short_acf)[2]`.

8.  Simulate with autocorrlelation set to 0

```{r}
AR1 <- arima.sim(list(ar=c(0.0)),10000)
par(mfrow=c(1,3))
plot(1:10000,AR1,type="l")
ar1_acf <- acf(AR1,main="ACF")
ar1_pacf <-pacf(AR1,main="PACF")
```

## Exercise 2

