---
title: "Homework 3"
author: "Richard Palmer"
date: "9/11/2019"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(car)
library(lsmeans)
```

## Conceptual Questions

### Necessary Assumptions for Two Way ANOVA

- The populations are approximately normal
- Independence of Samples
- Equal variance among populations
- Groups must have the same sample size

### Interaction between two explanatory variables means..

That the nature of the effect one explanatory variable has on
the dependent variable is dependent on the value of the other 
explanatory variable.

### The family wise error rate is ..

The probability of making at least one Type 1 error (incorrectly rejecting the null
hypothesis when it is in fact true) among all the t-test comparisons that are 
done in an analysis.

It is an issue for ANOVA because ANOVA implicitly includes many sample comparisons.
The probability of making at least one error grows with the number of 
comparions that are made.

### The overall type-III sums of squares F-test's allow the analysis to determine
where specific differences lie between levels.

FALSE.  The F-test's will only indicate which factors are statically significant;
They will say nothing about the particular levels of any factor.

## Exercises

### ACT scores revisited

```{r summaries}
ACT<-read.csv("../Unit3PreLive/MathACT.csv")
#Attaching the data set, creating a function, and creating a summary stats table.  Note: In line 44 below, you can add other statistics like median, IQR,etc.
attach(ACT)
mysummary<-function(x){
  result<-c(length(x),mean(x),sd(x),sd(x)/length(x),min(x),max(x),IQR(x))
  names(result)<-c("N","Mean","SD","SE","Min","Max","IQR")
  return(result)
}
sumstats<-aggregate(Score~Background*Sex,data=ACT,mysummary)
sumstats<-cbind(sumstats[,1:2],sumstats[,-(1:2)])
sumstats
```

Standard Error plot
```{r}
ggplot(sumstats,aes(x=Background,y=Mean,group=Sex,colour=Sex))+
  ylab("ACT Score")+
  geom_line()+
  geom_point()+
  geom_errorbar(aes(ymin=Mean-SE,ymax=Mean+SE),width=.1)+
  ggtitle("Original SE-based error plot")
```

Error (raw SD) plot
```{r}
ggplot(sumstats,aes(x=Background,y=Mean,group=Sex,colour=Sex))+
  ylab("ACT Score")+
  geom_line()+
  geom_point()+
  geom_errorbar(aes(ymin=Mean-SD,ymax=Mean+SD),width=.1)+
  ggtitle("New SD-based error plot")
```

The SD-based error plot is more indicative of the actual data variances,
since it is not modified with the size of the data sample.

### Conducting a Two Way ANOVA analysis in R

Fit the model 
```{r modelfit}
model.fit<-aov(Score~Background+Sex+Background:Sex,data=ACT)
```

Use Tukey HSD to make all pairwise comparisons
```{r tukey }
TukeyHSD(model.fit,"Background:Sex",conf.level=.95)
```

What contrasts are we interested in?
```{r}
contrast.factor<-~Background*Sex
mycontrast<-c("amale-afemale","bmale-bfemale","cmale-cfemale")
dat<-ACT
```

Turner's magic code
```{r}
library(limma)
final.result<-c()
for( j in 1:length(mycontrast)){
contrast.factor.names<-gsub(" ", "", unlist(strsplit(as.character(contrast.factor),split = "*", fixed = T))[-1])
contrast.factor.2 <- vector("list", length(contrast.factor.names))
for (i in 1:length(contrast.factor.names)) {
  contrast.factor.2[[i]] <- levels(dat[, contrast.factor.names[i]])
}
new.factor.levels <- do.call(paste, c(do.call(expand.grid, 
                                              contrast.factor.2), sep = ""))
temp.cont<-mycontrast[j]
contrast2 <- list(comparison = as.vector(do.call(makeContrasts, 
                                                list(contrasts = temp.cont, levels = new.factor.levels))))
contrast.result <- summary(contrast(lsmeans(model.fit, 
                                            contrast.factor), contrast2, by = NULL))
final.result<-rbind(final.result,contrast.result)
}
#Cleaning up and applying bonferroni correction to the number
#of total comparisons investigated.
final.result$contrast<-mycontrast
final.result$bonf<-length(mycontrast)*final.result$p.value
final.result$bonf[final.result$bonf>1]<-1
final.result

```

The Tukey method produces for a:male to a:female a mean difference
estimate of 2.3851626.  Looking at the specific contrast of interest
also produces 2.3851626.  The p-value from Tukey HSD is 0.0854058,
while that from the contrats is 0.024165.  The difference can be
attributed to the larger family-wise error rate assumed by TukeyHSD,
which is making all possible pair-wise corrections.  When looking 
at three of the comparisons, there there are only three pair-wise
comparisons, so the adjustment is smaller.

### Iridium Depth

Get the data
```{r}
library(Sleuth3)
iridium_stats <- aggregate(Iridium~Strata*DepthCat,data=ex1317,mysummary)
iridium_stats <- cbind(iridium_stats[,1:2],iridium_stats[,-(1:2)])
iridium_stats
```

Do a means plot
```{r}
ggplot(iridium_stats,aes(x=DepthCat,y=Mean,group=Strata,colour=Strata))+
  ylab("Iridium Level")+
  geom_line()+
  geom_point()+
  geom_errorbar(aes(ymin=Mean-SE,ymax=Mean+SE),width=.1)
```



